Longitudinal Manifold Estimation

### 2023-04-25 paper/main.tex
add tuning section draft

We denote the PME coefficients at each time point $t$ as $\omega_i = [\mathbf{s}_t, \mathbf{\alpha}_t]$. To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the PME coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$. With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function
\[%
  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\phi(t)_j
,\]%
where, with matrix $\mathbf{A}$ defined as $\mathbf{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\mathbf{T}$ defined as $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = diag(w_1, \dots, w_T)$,  $\mathbf{\delta}$ and $\mathbf{\nu}$ are the solutions to 
\[%
  \left(
  \begin{array}{ccc}
    2\mathbf{A}\mathbf{W}\mathbf{A} + 2\gamma\mathbf{A} & 2\mathbf{A}\mathbf{W}\mathbf{T} & \mathbf{T} \\
    2\mathbf{T}^{T}\mathbf{W}\mathbf{A} & 2\mathbf{T}^{T}\mathbf{W}\mathbf{T} & \mathbf{0} \\
    \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \mathbf{\delta} \\
    \mathbf{\nu} \\
    \mathbf{m}
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\mathbf{A}\mathbf{W}\mathbf{\omega} \\
    2\mathbf{T}^{T}\mathbf{W}\mathbf{\omega} \\
    \mathbf{0}
  \end{array}
  \right)
.\]%

@@

@@

Contributions:

Limitations:
\begin{enumerate}
  \item Algorithm assumes stability in shape over time.
  \item Algorithm initialization is sensitive to magnitude of change between time points.
  \item Computation time.
\end{enumerate}

>>>>

We denote the manifold coefficients at each time point $t$ as $\omega_i = [\mathbf{s}_t, \mathbf{\alpha}_t]$. To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$. With a given tuning value $\gamma$, we then follow the estimation approach described in \cite{mengPrincipalManifoldEstimation2021} to estimate spline function
\[%
  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \eta_{4-d}\left(\|t - t_i\|\right) + \sum_{j=1}^{2}\nu_j\phi(t)_j
,\]%
where, with matrix $\mathbf{A}$ defined as $\mathbf{A}_{ij} = \eta_d\left(\|t_i - t_j\|\right)$, matrix $\mathbf{T}$ defined as $\mathbf{T}_{ij} = \phi_i(t_j)$, and $\mathbf{W} = diag(w_1, \dots, w_T)$,  $\mathbf{\delta}$ and $\mathbf{\nu}$ are the solutions to 
\[%
  \left(
  \begin{array}{ccc}
    2\mathbf{A}\mathbf{W}\mathbf{A} + 2\gamma\mathbf{A} & 2\mathbf{A}\mathbf{W}\mathbf{T} & \mathbf{T} \\
    2\mathbf{T}^{T}\mathbf{W}\mathbf{A} & 2\mathbf{T}^{T}\mathbf{W}\mathbf{T} & \mathbf{0} \\
    \mathbf{T}^{T} & \mathbf{0} & \mathbf{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \mathbf{\delta} \\
    \mathbf{\nu} \\
    \mathbf{m}
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\mathbf{A}\mathbf{W}\mathbf{\omega} \\
    2\mathbf{T}^{T}\mathbf{W}\mathbf{\omega} \\
    \mathbf{0}
  \end{array}
  \right)
.\]%

With $\gamma$ given, we denote the manifold coefficients at time $t$ estimated by function $g(\cdot)$ as $w_{\gamma}(t) = \left[\mathbf{s}_{\gamma}(t), \mathbf{\alpha}_{\gamma}(t)\right]$. Hence, given $\gamma$, the estimated embedding function given time $t$ and $d$-dimensional parameterization $\mathbf{r}$ may be written as
\[%
  f_{\gamma}(t, \mathbf{r}) = \sum_{j=1}^{N}\mathbf{s}_{\gamma}(t)_j \eta_{4-d}\left(\|\mathbf{r} - \mathbf{r}_j^*\|\right) + \sum_{k=1}^{d+1}\mathbf{\alpha}_{\gamma}(t)_k \phi_k(\mathbf{r})
.\]%

@@

Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross-validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
\[%
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2
.\]%

The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.

@@

\subsection*{Contributions}

\subsection*{Limitations}

\begin{enumerate}
  \item Algorithm assumes stability in shape over time.
  \item Algorithm initialization is sensitive to magnitude of change between time points.
  \item Computation time.
\end{enumerate}
